---
title: "EDA Final, Turtles Project"
author: "Eric Raju"
date: "10/30/2020"
output: pdf_document
---

```{r setup, include=FALSE}
options(knitr.duplicate.label = "allow")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(ggplot2)
```



## Introduction

This project involves the study of the ecology and survival of turtles in Algonquin park, Toronto, in a **mark-and-recapture experiment**. This essentially consists of arbitrarily selecting and tracking a sampling of some wild species and then sending them back into the wild, so to speak, and hopefully then being able to return at some time later to find them again. At each of these capture-recapture events, measurements are recorded, including the dimensions of the turtle and the ecological environment, so as to better assess an age estimate. Turtle ages in Algonquin Park, according to Dr. Niall, are correlated heavily with their size, and they seem to grow until adulthood at which point they eventually plateau.

Prior to any statistical analysis of survival trends or patterns, however, it is necessary to go through the data and see if any cleaning needs to be done. This generally consists of checking out outlying or missing values and trying to figure out the patterns that arise therefrom.

Variables and Notation Index:

* **notch**:	Unique individual ID based on permanent markings researchers made in individual shells
* **Turtle ID**:	A supplementary system of Iding each turtle, based on a tag put on each turtle's shell. There is no need to use this column

* **Unique site**:	Some code that depicts what sites it has een found at; I suggest making your own code if you need it from the site records, rather than relying on this.
known age	Whether or not a turtle has been followed since birth, or can otherwise be assigned an age. Not completely reliable.

* **Year of birth**:	Not reliable, but estimated year of birth (or known, from some known age individuals)

* **Year**:	Year that the row measurements we taken

* **age**:	Not reliable, but estimated year of birth (or known, from some known age individuals)

* **sex**:	Either UNKNOWN (immature individuals can't be sexed, unlike e.g., humans), or Male, or Female

* **juv**:	Logical, but not reliable for some reason. If you want to estimate whether an individual is a juvenile in a given year, then know that males mature around PL 9-11cm, females around PL 12-14cm. So any turtles PL< 9cm is a juvenile; any female PL< 12cm is a juvenile; "Unknown sex" individuals are usually juveniles

* **yearjuv**:	Ignored

* **date**:	The date of measurement

* **Julian date**:	The julian date of measurement

* **site**:	Only include these sites:

** Arowhon Rd.	Sims Creek	Maiden Lake	Road to WHP	Blanding's Bog	Wolf Howl Pond	Wolf Howl Pond E.	Road to Wrose	West Rose	March Hare Lake

* **repro**:	Ignored

* **gravid**:	Logical, if the turtle was found gravid (i.e., with eggs in her belly) in that year

* **nest**:	Logical, if we recorded one nest from that individual in that year

* **nestdate**:	Nest date

* **secnest**:	Logical, if a second nest was laid in a given year

* **nest2dat**:	Second nest date

* **firstcap**:	Whether or not this is the first time we have ever caught that turtle in the history of the project

* **mass**:	weight of the turtle 

* **cl**:	carapace length (maximum length)

* **pl**:	Plastron Length (maximum length)

* **midcl**:	Carapace length (Measured from between the anterior-most carapacial scutes to between the postior-most carapacial scutes)

* **midpl**:	Plastron length (Measured from between the anterior-most plastral scutes to between the postior-most plastral scutes)

* **lclaw**:	Length of the third nail of the left claw. Claw measurements were taken regularly up until about 2011 or so, then less frequently or not at all. Males develop very long claws as soon as they are sexually mature

* **rclaw**:	Length of the third nail of the right claw. Claw measurements were taken regularly up until about 2011 or so, then less frequently or not at all. Males develop very long claws as soon as they are sexually mature

* **cw**:	The width of the widest portion of the carapace

* **ch**:	The height of the turtle, measured from the bottom of the plastron to the top of the carapace.

* **dead**:	Wether the turtle was found dead

* **C**:	Irrelevant

* **P**:	Irrelevant

* **cannuli**:	Irrelevant

* **pannuli**:	Irrelevant

* **name**:	Name of the person collecting the morphological data

* **Permanent 9-11**:	Irrelevant

* **Permanent 11-13**:	Irrelevant

* **Permanent 13-15**:	Irrelevant

* **Permanent 15+**:	Irrelevant

* **VB age estimation**:	Irrelevant

* **C'**:	Irrelevant

* **P'**:	Irrelevant

## The Data Cleaning Process

The order of data cleaning I performed was, as follows:

1.	In Excel, I highlighted by the colour given all of the sites in
question that were to be analyzed.

2.	Furthermore, I deleted the columns which the investigator said
were superfluous, according to the notes in the description.

3.	I changed all of the dates to the time series format, YYYY-MM-DD, and transferred this data to R.

4.	I then considered the question of ID’ing the turtles. Notches were the primary method whereby we were supposed to identify them, so concentrating on these I tried to standardize these values. The problem is that some researchers may use different naming systems at different times, and some might be alphanumeric while others may be fully numeric. Among the alphanumeric ones, I noticed that all of them had 4 to 5 characters, whereas the maximum number of characters of the wholly numeric data was 4. Since notches of the type number-number-number-letter were generally of similar magnitude to the notches in the surrounding rows, I assumed that the remaining letters were unnecessary or a typo, and deleted them. Thus, with the exception of some notches containing the type “000X” (where X can be 0, A or B) or for some reason “DEAD”, all of them under a corrected notation had at most 4 wholly numeric characters. I added an extra column with a flag in each of these cases yielding TRUE when I had this “strange” data. I also deleted these latter in a “cleaned” version of the data when looking at plots, though further modelling should consider whether these could be typos (such as A = 4 or B = 8) and would thereby affect the data fidelity. One must also consider the distribution of these numbers to see whether if it follows traditional laws of randomness. Apparently, most of the “strange” data like this occurred in the site called “Wolf Howl Pond E.”

5.	Before going further I started looking at some patterns in the data, in particular over the time series. These patterns will be explored next.

6.	(I also noticed some of the earlier data, in the late 1980s, was very inconsistent and often contained missing data, so I also deleted the data for those years in the late 1980s regarding probable year of birth.)

7.	Lastly, I deleted all of the rows which contained any missing “cl”, “pl”, “midcl” or “midpl” data, i.e. columns corresponding to the crucial dimensions of the turtles’ size necessary to estimate age. This could obviously introduce bias.

A central question of survival research involves population and age estimation. One of the aforementioned principles included the method of allometry, using relative turtle size data to estimate overall age. While similar environmental conditions as would occur in a dense forest like Algonquin Park, it would still be wise to consider unexpected results or outliers in the estimation process. Thus, I looked at first of all, the distribution of the probable year of birth for a turtle sighted on any given year.

## Data Analysis

```{r tidyverse, stringr, ggplot2, readxl, echo=FALSE}

# Helper functions
cleanTable <- function(table) {
  # A function to transform a csv file into a table.
  clean.table <- c()
  for (row in 1:length(table)) {
    this.row <- strsplit(table[row],"\t")
    clean.table <- rbind(clean.table,as.character(this.row[[1]]))
  }
  clean.table
}

cleanChars <- function(table=clean.table) {
  # Convert all the numerical values to numeric type.
  n < nrow(clean.table)
  m <- ncol(clean.table)
  for (i in 1:n) {
    for (j in 1:m) {
      if (!(is.na(clean.table[i,j]))) {
        num <- as.numeric(clean.table[i,j])
        clean.table[i,j] <- num
      }
    }
  }
  clean.table
}

countUnique <- function(lst) {
  #Return each element in lst exactly once
  unique <- c()
  for (item in lst) {
    if (!(item%in%unique)) {
      unique <- append(unique,item,length(unique))
    }
  }
  unique
}

#index <- function(lst,item,k,all=FALSE) {
#  # Return the first index of item in lst. If k = -1, return the last one. Return
#  # only the one index is all is false, else return all.
#  indices <- c()
#  n <- length(lst)
#  for (i in 1:n) {
#    if (item==lst[i]) {
#      indices <- append(indices,i,length(indices))
#    }
#  }
#  if ((k!=-1)&&(!all)) {
#    indices[k]
#  }
#  else if (k==-1) {
#    n <- length(indices)
#    indices[n]
#  }
#  else {
#    indices
#  }
#}

###############

turtles_site <- readxl::read_excel("turtles_site.xlsx") # I copied directly from excel
my.file <- turtles_site
my.file <- as.data.frame(my.file)

notch.corrected <- my.file$notch
n <- 9973
notch.flags <- rep(FALSE,n)
for (i in 1:n) {
  any <- FALSE
  for (letter in c("A","B","C","D","E","F","G","H","I","J","K","L","M","N","O","P","Q","R","S",
                   "T","U","V","W","X","Y","Z")) {
    if (stringr::str_count(my.file$notch[i],letter)>=1) {
      any <- TRUE
    }
  }
  if (nchar(as.character(my.file$notch[i]))==5) {
    temp <- as.character(notch.corrected[i])
    notch.corrected[i] <- substr(temp,start=1,stop=nchar(temp)-1) # remove the letter
  }
  notch.flags[i] <- any
}

sites <- countUnique(my.file$site)
goodsites <- c("Arowhon Rd.","Sims Creek","Maiden Lake","Road to WHP","Blanding's Bog",
"Wolf Howl Pond","Wolf Howl Pond E.","Road to Wrose","West Rose","March Hare Lake")

# Strange indices and naming
my.file$'notch-correction' <- notch.corrected
# my.file$notch.flags <- notch.flags

#####

notch.trueInds <- c()
for (i in 1:n) {
  if (my.file$notch%in%my.file$my.file$'notch-correction') {
    notch.trueInds <- append(notch.trueInds,i,length(notch.trueInds))
  }
}
strange1 <- my.file[notch.trueInds,]
strange1 <- rbind(strange1[1,],strange1[4,])
nrow1 <- nrow(strange1)
str.rows1 <- notch.trueInds
 
strange1.notches <- vector("list",10)
strange1.good <- c()
for (i in 1:n) {
  if (strange1$site[i]%in%goodsites) {
    strange1.good <- append(strange1.good,i,length(strange1.good))
  }
}

for (i in 1:length(goodsites)) {
  this.notch <- strange1$`notch-correction`[strange1.good]
  strange1.notches[i][[1]] <- this.notch
}
 
strange1.notch2 <- vector("list",6)
for (j in 1:8) {
  if (j%in%c(1,2,3)) {
    strange1.notch2[j][[1]] <- strange1.notches[j][[1]]
}
  else if (j==4) {
    strange1.notch2[4][[1]] <- c(strange1.notches[4][[1]],strange1.notches[5][[1]])
}
  else if (j==6) {
    strange1.notch2[5][[1]] <- c(strange1.notches[6][[1]],strange1.notches[7][[1]])
}
   else if (j==8) {
     strange1.notch2[6][[1]] <- c(strange1.notches[8][[1]],strange1.notches[9][[1]],strange1.notches[10][[1]])
     }
}
 
#####

correct <- my.file$'notch-correction'

correct.na.out <- c()
correct.na.indices <- c()
for (j in 1:n) {
  if (!(is.na(as.numeric(as.character(correct[j]))))) {
    correct.na.out <- append(correct.na.out,as.character(correct[j]),length(correct.na.out))
    correct.na.indices <- append(correct.na.indices,j,length(correct.na.indices))
  }
}

strange2 <- my.file[correct.na.indices,]
nrow2 <- nrow(strange2)
row0 <- c()
for (i in 1:n) {
  if (my.file$notch[i]=="0000") {
    row0 <- append(row0,i,length(row0))
  }
}
str.rows2 <- row0

strange2.notches <- vector("list",10)
strange2.good <- c()
for (i in 1:n) {
  if (strange2$site[i]%in%goodsites) {
    strange2.good <- append(strange2.good,i,length(strange2.good))
  }
}
this.notch <- c()
for (i in 1:length(goodsites)) {
  this.notch <- strange2$`notch-correction`[strange2.good]
  strange2.notches[i][[1]] <- this.notch
}

strange2.notch2 <- vector("list",6)
for (j in 1:8) {
   if (j%in%c(1,2,3)) {
     strange2.notch2[j][[1]] <- strange2.notches[j][[1]]
   }
   else if (j==4) {
     strange2.notch2[4][[1]] <- c(strange2.notches[4][[1]],strange2.notches[5][[1]])
   }
   else if (j==6) {
     strange2.notch2[5][[1]] <- c(strange2.notches[6][[1]],strange2.notches[7][[1]])
   }
   else if (j==8) {
     strange2.notch2[6][[1]] <- c(strange2.notches[8][[1]],strange2.notches[9][[1]],strange2.notches[10][[1]])
   }
}

#####

dead.true <- c()
for (i in 1:n) {
  if (my.file$dead[i]==TRUE) {
    dead.true <- append(dead.true,i,length(dead.true))
  }
}
stranged <- my.file[dead.true,]
nrowd <- nrow(stranged)
str.rowsd <- dead.true

stranged.notches <- vector("list",10)
stranged.good <- c()
for (i in 1:n) {
  if (stranged$site[i]%in%goodsites) {
    stranged.good <- append(stranged.good,i,length(stranged.good))
  }
}
for (i in 1:length(goodsites)) {
   this.notch <- stranged$`notch-correction`[stranged.good]
   stranged.notches[i][[1]] <- this.notch
}

stranged.notch2 <- vector("list",6)
for (j in 1:8) {
   if (j%in%c(1,2,3)) {
     stranged.notch2[j][[1]] <- stranged.notches[j][[1]]
   }
   else if (j==4) {
     stranged.notch2[4][[1]] <- c(stranged.notches[4][[1]],stranged.notches[5][[1]])
   }
   else if (j==6) {
     stranged.notch2[5][[1]] <- c(stranged.notches[6][[1]],stranged.notches[7][[1]])
   }
   else if (j==8) {
     stranged.notch2[6][[1]] <- c(stranged.notches[8][[1]],stranged.notches[9][[1]],stranged.notches[10][[1]])
   }
}
 
#####
 
good <- my.file[-c(str.rows1,str.rows2,str.rowsd),] # the "good" data
 
good.notches <- list(rep(0,10))
good.correct <- c()
for (i in 1:n) {
  if (good$site[i]%in%goodsites) {
    good.correct <- append(good.correct,i,length(good.correct))
  }
}

for (i in 1:length(goodsites)) {
  this.notch <- good$`notch-correction`[good.correct] 
  good.notches[i][[1]] <- this.notch
}

good.notch2 <- vector("list",6)
for (j in 1:8) {
  if (j%in%c(1,2,3)) {
    good.notch2[j][[1]] <- good.notches[j][[1]]
  }
  else if (j==4) {
    good.notch2[4][[1]] <- c(good.notches[4][[1]],good.notches[5][[1]])
  }
  else if (j==6) {
    good.notch2[5][[1]] <- c(good.notches[6][[1]],good.notches[7][[1]])
  }
  else if (j==8) {
    good.notch2[6][[1]] <- c(good.notches[8][[1]],good.notches[9][[1]],good.notches[10][[1]])
  }
}
 
# flag.logic <- c()
# nall <- n
# for (row in 1:nall) {
#   flag.logic <- append(flag.logic,identity(my.file$notch.flags[row],val=TRUE),length(flag.logic))
# }
 
good.nums <- c()
good.k <- c()
for (k in 1:6) {
  good.nums <- append(good.nums,good.notch2[k][[1]],length(good.nums))
  good.k <- append(good.k,k,length(good.k))
}
good.nums <- as.numeric(good.nums)
 good.all <- my.file[good.k]
strange.all <- my.file[-good.k]

# Editing:
for (k in 1:3) {
  strange1.notch2[k][[1]] <- 0
  strange2.notch2[k][[1]] <- 0
  stranged.notch2[k][[1]] <- 0
}
strange1.notch2[4][[1]] <- "OOOA"
strange2.notch2[4][[1]] <- rep("0000",4)
stranged.notch2[4][[1]] <- NA
for (k in 5:6) {
  strange1.notch2[k][[1]] <- strange1.notch2[k][[1]][1]
  stranged.notch2[k][[1]] <- stranged.notch2[k][[1]][1]
}
strange2.notch2[5][[1]] <- rep("0000",10)
strange2.notch2[6][[1]] <- NA

sites.matrix <- matrix(0,ncol=6,nrow=4)
colnames(sites.matrix) <- c(goodsites[1:3],paste(goodsites[4],goodsites[5]),
                                paste(goodsites[6],goodsites[7]),
                                paste(goodsites[8],goodsites[9],goodsites[10]))
rownames(sites.matrix) <- c("Good","OOOA or OOOB","OOOO","DEAD")
for (snum in 1:6) {
  sites.matrix[1,snum] <- length(good.notch2[snum][[1]])
  scond <- identity(is.na(strange2.notch2[snum][[1]][1]))
  sites.matrix[2,snum] <- length(strange1.notch2[snum][[1]])*scond
  sites.matrix[3,snum] <- length(strange2.notch2[snum][[1]])*scond
  sites.matrix[4,snum] <- length(stranged.notch2[snum][[1]])*scond
}

typeofstrange <- sites.matrix[1,]
colsum <- c()
for (snum in 1:6) {
   this.col <- sites.matrix[2,snum]+sites.matrix[3,snum]+sites.matrix[4,snum]
   colsum <- append(colsum,this.col,length(colsum))
 }
typeofstrange <- rbind(typeofstrange,colsum)
typeofstrange <- rbind(typeofstrange,(typeofstrange[2,]/typeofstrange[1,]))
colnames(typeofstrange) <- colnames(sites.matrix)
rownames(typeofstrange) <- c("Good","Strange","Fraction strange")

fract <- table(typeofstrange[3,])

# Plotting

# Where are the weird data?
```

```{r pressure, tidyverse, stringr, ggplot2, echo=FALSE}
shortsites <- colnames(typeofstrange)
strange.data <- as.data.frame(t(typeofstrange))
# 
ggplot(data=strange.data,aes(x=shortsites,y=`Fraction strange`))+geom_point()+ggtitle("Weird cases by site")
```

Most of the weird cases with alpha-numeric id's come from "the road to WHP" or Blanding's Bog.

Probability of first capture by year--was there a bias?

``` {r code-chunk 1}
ggplot(data=my.file,aes(x=`Year of birth`,y=firstcap))+geom_point()+geom_line()+ggtitle("First capture vs. year of birth")
```

It turns out while barely slightly more younger turtles captured for the first time, despite the missing values there are many more older turtles which were not captured the first time.

How many were captured for a given date with a certain year of birth?

``` {r code-chunk 2}
all.years <- sort(countUnique(my.file$Year))
year.to.birth <- c()
var.birth <- c()
for (y in 1:length(all.years)) {
  year.file <- c()
  temp <- c()
  for (z in 1:y) {
    year.file <- append(year.file,z,length(year.file))
  }
  this.years <- my.file$`Year of birth`[year.file]
  temp <- this.years
  year.data <- c()
  for (item in temp) {
    if (!(is.na(item))) {
      year.data <- append(year.data,item,length(year.data))
    }
  }
  year.mean <- mean(year.data)
  this.birth <- max(year.data)-min(year.data)
  year.to.birth <- append(year.to.birth,year.mean,length(year.to.birth))
  var.birth <- append(var.birth,this.birth,length(var.birth))
}
```

Note, the value for 1989 has NA so use the previous years' last entry.

``` {r code-chunk 3}
yearbirth <- my.file[,3]
ageknown <- my.file[,2]
my.file$yearbirth <- as.character(yearbirth)
my.file$ageknown <- as.character(ageknown)
colnames(my.file)[30] <- "yearbirth"
colnames(my.file)[31] <- "ageknown"

all.years <- as.numeric(all.years)
year.to.birth <- as.numeric(year.to.birth)
births <- cbind(all.years,year.to.birth)
births <- as.data.frame(births)

birthlm <- lm(year.to.birth~all.years,data=births)
plot(all.years,year.to.birth)

values <- vector("list",32)
for (i in 1:32) {
 if (!(i==4)) {
   values[i][[1]] <- fitted(birthlm)[i][[1]]
 }
 else {
   values[i][[1]] <- values[i-1][[1]]
 }
}
values2 <- c()
for (i in 1:32) {
  values2 <- append(values2,values[i][[1]],length(values2))
}
values2[32] <- values2[31]

lines(all.years,values2,col="red")
```

We can more clearly plot the birth ranges per year:
``` {r code-chunk 4}
plot(all.years,year.to.birth)
```

There's a clear linear trend, though with some deviation in the earlier period of 1980-1990. It seems that the line of best fit implies that the number of years from the birth of the observed turtle is proportional to the year observed. However, there appears to be an increasing frequency over time (some of the observations even seem to oscillate and get closer and closer to the line). It appears that we cannot assume that the count of observations are independent of time.

Going back to the original year of birth vs. age plot: is the age estimate good?

``` {r code-chunk 5}
ggplot(data=my.file,aes(x=yearbirth,y=age))+geom_point()+geom_line()+
  ggtitle("Age vs. year of birth, known vs. unknown age")

# Comparing known versus unknowns:
known.turtles <- c()
for (i in 1:n) {
  if (my.file$ageknown[i]==TRUE) {
    known.turtles <- append(known.turtles,i,length(known.turtles))
  }
}
known <- my.file[known.turtles,]
unknown <- my.file[-known.turtles,]

ggplot(data=known,aes(x=`Year of birth`,y=age))+geom_point(col="green")+geom_line(col="green")+
  geom_point(data=unknown,aes(x=`Year of birth`,y=age),col="yellow")+
  geom_line(data=unknown,aes(x=`Year of birth`,y=age),col="yellow")+ggtitle("Age vs. year of birth, partitioned")
```

It seems that, with some outliers, the predicted trend generally follows the values without assuming whether they are known or unknown. However, it seems to be less exact for the data before the observations of those turtles born around the late 1980s.

What are the sources of missing values?

``` {r code-chunk 6}
nas <- c()
for (col in 1:ncol(my.file)) {
  nas <- append(nas,sum(is.na(my.file[,col]),length(nas)))
}
```

Notes:
* The two columns with the largest number of missing values, both over 5000, are the year of first and second nest dates.

* The second nest date column is emptier because by definition one needs a first nest date beforehand.

* The next largest NA columns are those corresponding to year of birth and age, with both over 3000 counts.

* The other columns with significant (i.e. over 100) missing data are the columns 18-28 which contain measurements.

To see the impact of missing values, consider the year of birth and age versus first and second nest date.

So... what to do with the missing measurement data? Most of the missing values occurred in the first four data points, so perhaps we want to eliminate everything that occured during this period.

``` {r code-chunk 7}
clean.file <- my.file[-which(my.file$Year%in%c(1978,1984,1985,1989)),]
```

Furthermore, we should eliminate any data with missing cl, pl, midcl or midpl as this will bias any estimate to age. Consider which years, sites and researchers are more likely to have this missing data.

``` {r code-chunk 8}
clean.file <- my.file[-which(my.file$Year%in%c(1978,1984,1985,1989)),]
clpl.out <- c()
bad.clpl <- c()
for (row in 1:n) {
  if (!((is.na(clean.file$cl[row]))||(is.na(clean.file$pl[row]))||
        (is.na(clean.file$midcl[row]))||(is.na(clean.file$midpl[row])))) {
    clpl.out <- rbind(clpl.out,my.file[row,])
  }
  else {
    bad.clpl <- rbind(bad.clpl,my.file[row,])
  }
}

clean.file <- clpl.out

ggplot(data=bad.clpl,aes(site))+geom_histogram(stat="count")+
  ggtitle("Missing measurements, preventing a good age estimate")
```

Most missing measurements occur in Wolf Howl Pond.

What does this do to the probability of capture?

``` {r code-chunk 9}
ggplot(data=clean.file,aes(x=`Year of birth`,y=firstcap))+geom_point()+geom_line()+
  ggtitle("First capture vs. year of birth, revised")
```

The probability of capture becomes somewhat more skewed to the right, looking like a more reasonable dataset that can be modelled logistically without much error.

``` {r code-chunk 10}
ggplot(data=known,aes(x=`Year of birth`,y=nestdate))+geom_point(col="green")+geom_line(col="green")+
  geom_point(data=unknown,aes(x=`Year of birth`,y=nestdate),col="yellow")+
  geom_line(data=unknown,aes(x=`Year of birth`,y=nestdate),col="yellow")+ggtitle("First nest date vs. year of birth")
```

The first nest date is proportional to the year of birth, so the data here seems to be collected well, but there is some spotty data near more recent values after the mid-1990s.


``` {r code-chunk 11}
ggplot(data=known,aes(x=`Year of birth`,y=nest2dat))+geom_point(col="green")+geom_line(col="green")+
  geom_point(data=unknown,aes(x=`Year of birth`,y=nest2dat),col="yellow")+
  geom_line(data=unknown,aes(x=`Year of birth`,y=nest2dat),col="yellow")+ggtitle("Second nest date vs. year of birth")
```

This has a similar shape to the previous graph, though it is even more spotty. While most of this can be attributed to the fact that not all of the turtles will be able to lay more than one nest, there is some un-evenness in the sparsity suggesting bias due to missing data, especially for later dates of nestings.

# The measurements:

These are plots of various measurements to see what the trends are and if they are linear, to get a good idea of how accurate the estimates are.

``` {r code-chunk 12}
ggplot(data=clean.file,aes(x=Year,y=mass))+geom_point()+geom_line()+ggtitle("Mass vs. year")
ggplot(data=clean.file,aes(x=Year,y=cl))+geom_point()+geom_line()+ggtitle("Cl vs. year")
ggplot(data=clean.file,aes(x=Year,y=pl))+geom_point()+geom_line()+ggtitle("Midcl vs. year")
ggplot(data=clean.file,aes(x=Year,y=midcl))+geom_point()+geom_line()+ggtitle("Pl vs. year")
ggplot(data=clean.file,aes(x=Year,y=midpl))+geom_point()+geom_line()+ggtitle("Midpl vs. year")
ggplot(data=clean.file,aes(x=cl,y=midcl))+geom_point()+geom_line()+ggtitle("Cl vs. midcl")
ggplot(data=clean.file,aes(x=pl,y=midpl))+geom_point()+geom_line()+ggtitle("Pl vs. midpl")
ggplot(data=my.file,aes(x=cl,y=pl))+geom_point()+geom_line()+ggtitle("Cl vs. pl")
ggplot(data=clean.file,aes(x=Year,y=lclaw))+geom_point(col="red")+geom_line(col="red")+ggtitle("Lclaw vs. year")+
  geom_point(data=clean.file,aes(x=Year,y=rclaw),col="blue")+geom_line(data=my.file,aes(x=Year,y=rclaw),col="blue")+
  ggtitle("Rclaw vs. year")
ggplot(data=clean.file,aes(x=Year,y=cw))+geom_point()+geom_line()+ggtitle("Cw vs. year")
ggplot(data=clean.file,aes(x=Year,y=ch))+geom_point()+geom_line()+ggtitle("Ch vs. year")
ggplot(data=clean.file,aes(x=Year,y=dead))+geom_point()+geom_line()+ggtitle("Deaths vs. year")
ggplot(data=clean.file,aes(x=name,y=site))+geom_point()+geom_line()+ggtitle("Researchers by site")

deathssite <- c()
found <- c()
for (this.site in goodsites) {
  site.deaths <- 0
  for (i in 1:n) {
    if ((my.file$site[i]==this.site)&&(!(i%in%found))) {
      site.deaths <- site.deaths + 1
      found <- append(found,i,length(found))
    }
  }
  deathssite <- append(deathssite,site.deaths,length(deathssite))
}

ggplot(data=NULL,aes(deathssite))+geom_bar()+ggtitle("Deaths by site")
```

Let's go back to the estimates of age.

``` {r code-chunk 13}
known2 <- clean.file[known.turtles,]
unknown2 <- clean.file[-known.turtles,]

ggplot(data=known2,aes(x=`Year of birth`,y=nestdate))+geom_point(col="green")+geom_line(col="green")+
  geom_point(data=unknown2,aes(x=`Year of birth`,y=nestdate),col="yellow")+
  geom_line(data=unknown2,aes(x=`Year of birth`,y=nestdate),col="yellow")+
  ggtitle("First nest date vs. year of birth, revised")

ggplot(data=known2,aes(x=`Year of birth`,y=nest2dat))+geom_point(col="green")+geom_line(col="green")+
  geom_point(data=unknown2,aes(x=`Year of birth`,y=nest2dat),col="yellow")+
  geom_line(data=unknown2,aes(x=`Year of birth`,y=nest2dat),col="yellow")+ggtitle("Second nest date vs. year of birth")
```

The data is much cleaner and more even now.

## Next Steps

After this analysis I feel that it is necessary to look more into the reasons for the sparse and missing data, in particular, the relationship between year, age and nesting. This may influence for example counts of turtle populations, which can then be looked at through generation analyis.

We should also be comparing the numbers and ages of turtles within the specific areas of observation in the park, as some of the earlier exploratory plots have shown some heterogeneity in the data collection at the various sites.

Finally we would then perform a full analysis of the survival rates and information from the observations and estimated or known age to get a good idea of turtle population structure over time and how it is informed by the age distribution.